
---
title: "Energy-Transitions"
pagetitle: "Energy-Transitions — Class 2"
author: "Renan Magalhães"
output: bookdown::bs4_book
---



# Session 2 {-}

***
## 1. Why Data Cleaning Is the Hardest (and Most Important) Step{-}

<iframe width="720" height="405" src="https://www.youtube.com/embed/dQw4w9WgXcQ" title="Class 02 Recording" frameborder="0" allowfullscreen></iframe>

When researchers talk about “analysis,” they often imagine fancy models and colorful plots.
In reality, most of the work happens before any model is fitted: cleaning, organizing, and validating the raw data. Raw data are almost never ready for analysis. Surveys come back with extra spaces, numbers formatted as text, spelling variations, and missing responses.
If you skip the cleaning stage, every later step—summary statistics, regression, visualization—will quietly give misleading results.

**What “tidy” means**

Hadley Wickham defines tidy data as having three rules:

*1.*Each variable forms a column.
*2.*Each observation forms a row.
*3.*Each value forms a single cell.

Once data satisfy these rules, they can easily be filtered, summarized, plotted, and joined.
Most R packages (especially those in the tidyverse) assume data are tidy.
The cleaning workflow we’ll follow mirrors the professional pipeline:



```{r echo = FALSE}

library(gt)

workflow <- data.frame(
  Step = c("Import", "Inspect", "Clean names/types", "Handle missingness",
           "Recode", "Verify visually", "Save"),
  Goal = c("Load raw data reliably",
           "Understand structure",
           "Make columns usable",
           "Make absence explicit",
           "Create consistent categories",
           "Sanity-check patterns",
           "Freeze a clean version"),
  Example_Functions = c("read_csv()",
                        "glimpse(), summary()",
                        "janitor::clean_names(), mutate()",
                        "is.na(), if_else()",
                        "case_when(), fct_recode()",
                        "ggplot()",
                        "write_csv()")
)

workflow |>
  gt() |>
  tab_header(
    title = "Data Cleaning Workflow in R"
  ) |>
  cols_label(
    Step = "Step",
    Goal = "Goal",
    Example_Functions = "Example Functions"
  ) |>
  fmt_markdown(columns = everything())
```


## 2. Importing and Inspecting the Dataset{-}

We’ll use the same FinSESCo survey introduced in Session 1.
This time we’ll pay close attention to its structure, not yet to its content.


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
finsesco <- read_csv(here::here("data", "finsesco_dirt.csv"))
```


**How R reads files**

`read_csv()` (from readr) is faster and safer than base R’s `read.csv()`:

- It assumes UTF-8 encoding (so German umlauts, € signs, etc. remain intact).
- It guesses data types by scanning a sample of rows.
- It reports parsing problems instead of silently converting to text.

`here::here()` constructs file paths relative to your R Project root, which makes your script portable between computers and operating systems.


Now look inside:

```{r}
glimpse(finsesco)
summary(finsesco)
```

`glimpse()` prints one line per variable, showing type and first values; `summary()` gives numeric summaries or category counts.


Rows: 2,585
Columns: 45
$ Age                <chr> "32", "45", "63", "28", …
$ Income Range (€)   <chr> "2,500–4,000", "4,000–6,000", …
$ Gender             <chr> "Female", "Male", …


You can already you spot issues:

- `Age` is stored as character, not numeric.
- Column !!!!! contain spaces and symbols.
- There are multiple inconsistent formats for `income`.

Before cleaning, it helps to know what R thinks each type means:

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(gt)
library(tibble)

types <- tibble(
  `R Type` = c("character", "numeric", "integer", "factor", "logical"),
  `What It Represents` = c("text", "decimal numbers", "whole numbers", "categorical labels", "TRUE/FALSE flags"),
  `Common Problems` = c("numbers read as text", "missing decimals → integers", "rarely a problem", "wrong order or encoding", "blanks become NA")
)

types |>
  gt() |>
  tab_header(title = "Overview of R Data Types")
```



Check data types with `sapply(finsesco, class)` if you’re unsure.

::: {.callout-tip}

💡 Tip {-}

R’s type system is strict: `"42"` (character) is not the same as `42` (numeric).
Whenever arithmetic fails, first check the variable’s type.
:::


## 3. Cleaning Column Names and Correcting Types{-}


**Why clean names?**

Messy names make scripts fragile. Every time you use spaces or special characters, you have to surround the name with backticks, e.g. `Income Range (€)`.
`janitor::clean_names()` automatically fixes this by converting all names to lower_snake_case and stripping illegal characters.



```{r  message=FALSE, warning=FALSE}
library(janitor)

finsesco <- finsesco %>%
  clean_names() %>%
  mutate(age = as.integer(age),income_range = as.factor(income_range))

```


**Step-by-step explanation**

- `clean_names()` renames columns like `Income Range (€)` → `income_range_eur`. It also ensures uniqueness—if two columns share the same label, it appends _2, _3, etc.
- `mutate()` lets us change columns in-place.
- `as.integer(age)` converts character digits into true numbers.
- `as.factor(income_range)` tells R this is a categorical variable.

Re-inspect:


```{r}
glimpse(finsesco)
```

If you now see age: int and income_range: fct, the cleaning worked.

::: {.callout-important}

🔴 Important {-}

`as.numeric()` and `as.integer()` silently turn non-convertible text into `NA`.
Always check how many missing values you introduced: `sum(is.na(finsesco$age))`

:::


## 4. The Logic of Pipes and dplyr Verbs{-}

<iframe width="720" height="405" src="https://www.youtube.com/embed/dQw4w9WgXcQ" title="Class 02 Recording" frameborder="0" allowfullscreen></iframe>

The pipe operator `%>%` passes the output of one function as the first argument of the next. This creates readable “sentences” of code.

Example without pipes:

`summarise(group_by(finsesco, income_range),
          mean(index_1, na.rm = TRUE))`


Example with pipes:

```{r}
finsesco %>%
  group_by(income_range) %>%
  summarise(mean_index_1 = mean(index_1, na.rm = TRUE))
```

The second version mirrors human thought: take data → group → summarise.

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(gt)
library(tibble)

verbs <- tibble(
  Verb = c(
    "select()",
    "filter()",
    "mutate()",
    "arrange()",
    "summarise() + group_by()"
  ),
  `What it does` = c(
    "choose columns",
    "choose rows",
    "add/modify columns",
    "reorder rows",
    "aggregate groups"
  ),
  Analogy = c(
    "decide which questions to keep",
    "choose which participants to include",
    "compute new variables",
    "sort by value",
    "compute averages per category"
  )
)

verbs |>
  gt() |>
  tab_header(title = "The Five Essential dplyr Verbs") |>
  cols_label(
    Verb = "Verb",
    `What it does` = "What It Does",
    Analogy = "Analogy"
  )
```



Let’s apply them:


```{r}
mini <- finsesco %>%
  select(age, gender, income_range, index_1) %>%
  filter(!is.na(index_1)) %>%
  mutate(high_invest_intent = index_1 >= quantile(index_1, 0.75, na.rm = TRUE)) %>%
  arrange(desc(index_1))
```

Explanation:

- `select()` keeps only relevant variables.
- `filter(!is.na(index_1))` removes incomplete responses.
- `quantile(index_1, 0.75)` finds the top 25 % cutoff.
- `mutate()` creates a logical variable (TRUE/FALSE) indicating whether each person is in that top quartile.
- `arrange()` orders the dataset so the highest scores appear first.

```{r}
head(mini)
```

Now you can see the structure: a concise, consistent table for further analysis.


If you want to know all the other verbs from `dplyr`, you can download a *cheat sheet*, that serves as a manual for using it.

[📄 Download the *dplyr Cheat Sheet* (PDF)](https://media.datacamp.com/legacy/image/upload/v1660300796/Marketing/Blog/Manipulating_Data_in_dplyr_Cheat_Sheet.pdf)


## 5. Grouping, Summarizing, and Interpreting {-}

Aggregating data is the first step toward understanding patterns.


```{r}
mini %>%
  group_by(income_range) %>%
  summarise(n = n(), 
            mean_index_1 = mean(index_1, na.rm = TRUE),
            sd_index_1 = sd(index_1, na.rm = TRUE),
            share_high = mean(high_invest_intent))
```

- `group_by()` adds a “grouping” attribute without changing the data itself.
- `summarise()` collapses each group to one row by computing summary statistics.
- Because TRUE = 1 and FALSE = 0, `mean(high_invest_intent)` automatically yields the proportion of TRUEs.

Interpretation example:

Higher-income respondents show higher mean `index_1` and a larger share of high-intention investors, suggesting economic capacity plays a role in willingness to invest.

**Be careful:** these are descriptive patterns, not causal effects!
We haven’t yet controlled for age, ownership, education or other variables.



## 6. Recoding and Creating Ordered Factors {-}

Categorical variables should have intuitive labels and logical order.

Re-labeling income groups

```{r  message=FALSE, warning=FALSE}
finsesco <- finsesco %>%
  mutate(
  income_range = fct_recode(income_range,
  "Low" = "Under €1,500",
  "Middle" = "€1,500–€4,000",
  "High" = "Over €4,000") %>% 
  fct_relevel("Low","Middle","High")
)
```


- `fct_recode()` maps old labels to shorter ones.
- `fct_relevel()` defines the order used in plots and summaries.
- Without re-leveling, R might sort alphabetically, putting “Middle” before “Low”.

Creating age groups


```{r}
finsesco <- finsesco %>%
  mutate(
  age_group = case_when(
  age < 30 ~ "18–29",
  age < 45 ~ "30–44",
  age < 60 ~ "45–59",
  age >= 60 ~ "60+",
  TRUE ~ NA_character_) %>%
  factor(levels = c("18–29","30–44","45–59","60+"))
)
```

- `case_when()` checks each condition sequentially.
- The pipe into `factor()` turns the string labels into an ordered categorical variable, ensuring that 18–29 appears before 30–44 in graphs.

Why this matters: statistical models interpret factors based on their order—by defining it explicitly, you avoid misleading reference levels later.


## 7. Saving and Versioning Clean Data {-}

Never overwrite the raw file. Always save a new one that documents its lineage.

```{r}
clean <- finsesco %>%
  select(age, age_group, gender,
  income_range, index_1)
```


```{r}
write_csv(clean, here::here("data", "finsesco_clean.csv"))
```

This becomes the baseline for all later sessions.
If you later discover an error, you can regenerate the file from the same script rather than editing data manually.


## 8. Conceptual Reflection {-}

Cleaning data is not busywork—it’s modeling reality.
When you decide that “under €1,500” means “Low-Income” you’re interpreting social categories. When you replace a missing `index_1` by the median, you’re making an assumption about the underlying distribution. Recognizing these choices makes you a transparent analyst.



## Exercise {-}

- **Exercise 02**: **[Download PDF](exercises/ex02/ex02.Rmd)**  
- Upload via **Assignments** page on Moodle.





